---
title: "Making maps with R"
date: "`r Sys.Date()`"
author: "Alexandre Bevington, FNRORD"
output: md_document
---

```{r setup, include=F}
# knitr::opts_chunk$set(warning = F, echo = T, )

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "images/vignette-fig-",
  out.width = "100%",
  fig.width = 5,
  fig.height = 5,
  dev = "CairoPNG",
  dpi = 300
)
```

### Introduction

R is not a GIS like ArcMap or QGIS... it is generally harder to **zoom around and check our your data**. That said, there are a lot of advantages of using R for map making:

- **Open source**: sharing/collaborating
- **Reproducible**: science and decision making
- **Efficient**: modify/recreate code between projects
- **Data science**: integrate analysis into maps
- **Exploration**: many tools to get to know your data

Today we will run through three examples:

1. **Basics**: Very short intro!
1. **Example A**: Average maximum SWE
1. **Example B**: BEC units with a hillshade
1. **Example C**: Satellite images

Goal: 

- Showcase examples
- Share workflows 
- Summarise lessons learned

Audience: 

- **Beginers** for sample code and showcase
- **Intermediate** R users for technical interest
- **Advanced** Help me! 

***
### Basics
***

#### Step 1 - Load libraries

```{r libraries}
# .. getting data 
library(bcdata)     #access to bc data catalogue
library(bcmaps)     #handy for quick base layers (to be deprecated?)

# .. data manipulation and plotting 
library(ggplot2)    #plotting
library(dplyr)      #data wrangling
library(lubridate)  #helps with dates
library(patchwork)  #combine multiple plots in single figure
library(ggsn)       #add north arrow and scale bar

# .. geospatial 
library(sf)         #vector manipulations (see also `sp`)
library(raster)     #raster manipulations (see also `stars`, `terra`)
library(RStoolbox)  #useful remote sensing and raster tools

# .. labelling
library(ggrepel)    #label engine to avoid overlapping labels
```

#### Step 2 - ggplot basics

- The **GGPLOT()** 
  - This is the canvas!
- The **GEOM** 
  - These are the data layers
  - Ex: 
    - geom_point(), 
    - geom_line(), 
    - geom_sf(), 
    - geom_raster(), etc.
- The **AES()** 
  - These are the rules for how the data are plotted 
  - Map an aethetic to a variable aes(x = lon, y = lat, fill=group)
  - Ex: 
    - x = , 
    - y = , 
    - color = , 
    - fill = , 
    - shape = , etc.
- The **SCALE** 
  - This is how the aesthetics are represented
  - Ex: 
    - scale_color_binned, 
    - scale_x_continuous, 
    - scale_shape_manual, etc.
- The **THEME()** 
  - These allow you to format the map (Ex: format legend, aspect ratio, background etc.)

#### Step 3 - Let's make a map! 

```{r}
# Make Some Data (e.g. csv)
my_data <- data.frame(
  Name = "Prince George", 
  X = -122.709,
  Y = 53.910)

# Convert to "Simple Feature"
my_data_sf <- st_as_sf(my_data, coords = c("X","Y"), crs = 4326)

# Reproject
my_data_sf_albers <- my_data_sf %>% st_transform(3005)

# Buffer 
my_data_sf_albers_buf <- my_data_sf_albers %>% st_buffer(3000)


ggplot() +
  geom_sf(data = my_data_sf_albers_buf, 
          aes(fill = "30 km buffer (3005)")) +
  geom_sf(data = my_data_sf, 
          shape = 21, 
          fill = "black", 
          aes(size = "Prince George (4326)")) +
  scale_fill_manual(values = "red") +
  labs(fill = "Buffer", size = "Point", x = "Lon.", y = "Lat.") +
  coord_sf(crs = 3005) +
  theme_bw()

```

***
### Example 1: Maximum Snow 
***

#### Step 2 - Outline of BC

```{r outline of BC}
# Get outline of BC
bc <- bcmaps::bc_bound()

# Plot outline of BC
ggplot(bc) +
  geom_sf()

# Add a label to the map
ggplot(bc) +
  geom_sf() +
  geom_sf_label(label = "IS THIS BRITISH COLUMBIA?")

# Dissolve polygons to "multipolygon" and add label
map_bc <- bc %>%
  summarise() %>% # Dissolve
  ggplot() +
    geom_sf(fill = NA, color = "grey50", size = 0.5) +
    theme_bw() +
    labs(x = "Lon.", y = "Lat.", title = "Holy smokes!")

map_bc + geom_sf_label(label = "IS THIS BRITISH COLUMBIA?")
     # Map assigned to variable 
```

#### Step 3 - Where are the snow station stations?

```{r search_catalogue}
# Search catalogue
bcdc_search("snow pillows")

# Station locations
asws_loc <- bcdc_query_geodata("ebe546aa-ac34-491c-a828-fdc87fb70610", crs = 3005) %>%
  filter(STATUS == "Active") %>%
  collect()

# Plot locations
ggplot(asws_loc) +
  geom_sf(data = bc) + 
  geom_sf_label(aes(label = LOCATION_ID))
```

#### Step 4 - Where are the data?

```{r, fig.height = 4}
# Get the historical data
asws_swe_historical <- bcdc_get_data(record = '5e7acd31-b242-4f09-8a64-000af872d68f',
                                     resource = 'f4ec0b1f-f8ba-4601-8a11-cff6b6d988a4') %>%
  filter(variable == "Snow_Water_Equivalent")

# Inspect data at a random station
asws_swe_historical %>%
  filter(Pillow_ID == "1A14P") %>% 
  ggplot() + 
    geom_line(aes(Date, value)) + 
    labs(title =  "1A14P") + 
    theme(aspect.ratio = 0.4)

# Remove negative values
asws_swe_historical <- asws_swe_historical %>% filter(value >= 0)

# Inspect data again
asws_swe_historical %>%
  filter(Pillow_ID == "1A14P") %>% 
  ggplot() + 
    geom_point(aes(Date, value)) + 
    geom_line(aes(Date, value)) + 
    labs(title =  "1A14P") + 
    theme(aspect.ratio = 0.4)
```

#### Step 5 - Calculate mean annual maximum SWE and plot

```{r}  
# Mean annual maximum SWE
asws_swe_historical_mean <- asws_swe_historical %>% 
  mutate(year = year(Date)) %>% 
  # Calculate MAX and SD of SWE per station/year
  group_by(Pillow_ID, year) %>% 
  summarise(max_val = max(value, na.rm = T),
            sd_val = sd(value, na.rm = T)) %>% 
  # Calculate MEAN of the annual maximum and SD per station
  group_by(Pillow_ID) %>% 
  summarise(mean_max_val = mean(max_val, na.rm = T),
            mean_sd_val = mean(sd_val, na.rm = T))

# Join table to points
asws_slp <- dplyr::right_join(
  x = asws_loc, 
  y = asws_swe_historical_mean %>% mutate(LOCATION_ID = Pillow_ID))
```

#### Step 6 - Calculate mean annual maximum SWE and plot

```{r}
# Plot!
map_bc_swe <- map_bc +
  geom_sf(
    data = asws_slp, 
    aes(
      group = LOCATION_ID, 
      fill = mean_max_val, 
      size = mean_sd_val), 
    shape = 21) +
  scale_fill_distiller(
    palette = "Blues", 
    guide = guide_colorbar(frame.colour = "black", 
                           ticks.colour = "black", 
                           barheight = 8)) +
  labs(
    fill = "Max SWE (mm)", 
    size = "SD SWE (mm)", 
    title = "Max and SD of SWE in BC") + 
  geom_text_repel(
    data = asws_slp,
    aes(label = LOCATION_ID, geometry = geometry),
    stat = "sf_coordinates")

map_bc_swe
```

#### Step 7 - Optional - Mapview and Plotly

```{r, eval=F}
plotly::ggplotly(map_bc_swe)
mapview::mapview(asws_slp, zcol = "mean_max_val")
```

***
### Example 2 - BEC and hillshade
***

#### Step 1 - Read data

```{r}
# Read Data
bowron <- read_sf("data/bowron.sqlite")
hs <- raster("data/hs.tif")

# Initial Plot
ggR(hs) + geom_sf(data = bowron, fill = NA, color = "red")

# Import BEC data
bec <- bcmaps::bec() 

```
  
#### Step 2 - Read BEC, calculate area, plot

```{r, fig.height=13}

# Clip to Bowron
bec_bow <- st_intersection(bec, bowron) 

# Dissolve by Zone
bec_bow_dissolve <- bec_bow %>% 
  group_by(ZONE, ZONE_NAME) %>% 
  summarise()

# Calculate Area

bec_bow_dissolve$area_km2 <- round(units::set_units(st_area(bec_bow_dissolve), "km2"))

# Grab Colors
bec_col <- tibble(ZONE=names(bec_colors()),
                  BEC_COL=bec_colors())
  
# Join colors to polygons
bec_bow_dissolve <- left_join(bec_bow_dissolve, bec_col) 

# Make a map
bec_map <- ggR(hs) + 
  geom_sf(data = bec_bow_dissolve, 
          aes(fill = ZONE_NAME), 
          alpha = 0.5, 
          show.legend = F) +
  geom_sf(data = bowron, 
          fill = NA) + 
  scale_fill_manual(values = bec_bow_dissolve$BEC_COL) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +  
  labs(x = "", y = "", fill = "BEC Zone") 

# Make a bar plot
bec_stats <- ggplot(bec_bow_dissolve) +
  geom_col(aes(ZONE, as.numeric(area_km2), 
               fill = ZONE_NAME)) +
  geom_text(aes(ZONE, 
                as.numeric(area_km2), 
                label = as.numeric(area_km2)), vjust = 0, nudge_y = 50) +
  scale_y_continuous(limits = c(0, 1900), expand = c(0,0)) +
  scale_fill_manual(values = bec_bow_dissolve$BEC_COL) +
  theme_classic() +
  theme(aspect.ratio = 0.3,
        legend.position = "bottom", 
        legend.direction = "vertical") + 
  labs(y = bquote("Total area (km"^2~")"))

# Make a 2 panel map
bec_map/bec_stats

```

***
### Example 3 - Satellite Image
***

Sorry not enough time to develop new content! 
We will follow this tutorial: 

https://github.com/bcgov/bcgov-r-geo-workshop/blob/master/20191106_Day_2_PM_Raster/03_Raster_Sentinel.Rmd

### Outstanting topics 

1. Add scale bar and north arrow - see ggsn
1. Add WMS basemap - see ggmap
1. Save as: use ggsave()
